---
title: Time Series Data Analysis
author: 'Ilayda Dinc'
date: '2020-02-23'
slug: internal-project
categories: []
tags:
  - Academic
  - R Markdown
subtitle: 'Time Series Analysis of Steelhead Salmon Passage'
summary: 'Main objectives of this project are to examine the changes in the Steelhead Salmon (Oncorhynchus mykiss) passage across the Bonneville Dam in Oregon between the years 1940 and 2019, and to identify the seasonality and cyclycality of the passage if present. The data is provided by Columbia Basin Research [1] and it is analyzed to see whether there are seasonal patterns and/or bigger trends in salmon passage.'
authors: [Ilayda Dinc]
lastmod: '2020-03-01T15:02:39-08:00'
featured: no
image:
  caption: 'Steelhead Salmon (Oncorhynchus mykiss). Source:NOAA,https://www.fisheries.noaa.gov/species/steelhead-trout'
  focal_point: Smart
  preview_only: no
projects: []

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE,warning=FALSE)
```


```{r include=FALSE}
# load packages

library(tidyverse)
library(here)  
library(janitor) 
library(kableExtra)
library(lubridate)
library(paletteer)

# For ts stuff: 
library(tsibble)
library(fable)
library(fabletools)
library(feasts)
library(forecast)

```


Main objectives of this project are to examine the changes in the Steelhead Salmon (Oncorhynchus mykiss) passage across the Bonneville Dam in Oregon between the years 1940 and 2019, and to identify the seasonality and cyclycality of the passage if present. The data is provided by Columbia Basin Research [1] and it is analyzed to see whether there are seasonal patterns and/or bigger trends in salmon passage.


```{r message=FALSE, warning=FALSE, fig.align='center'}

#read in the data

salmon_data <- read_csv("cbr_fish.csv") %>%
  clean_names()

#unique(salmon_data$location)
#unique(salmon_data$parameter)
#summary(salmon_data)

# single location, single parameter

# arrange dates
salmon_date <- salmon_data %>%
  separate(mm_dd,into=c("day","month"),sep="-") %>% 
  mutate(yy_mm_dd=ymd(paste(year,month,day)))%>% 
    mutate(month_sep = yearmonth(yy_mm_dd)) %>%
  drop_na(month_sep)


#plotting daily passage

salmon_graph<-ggplot(data = salmon_date, aes(x = yy_mm_dd, y = value)) +
  geom_line(color='midnightblue') +
  theme_bw()+
  labs(x = "Year", y = "Daily Steelhead Salmon Passage")+
  theme(text = element_text(size=12))+
  scale_x_date(date_breaks = "5 year", date_labels = "%Y", limit=c(as.Date("1939-12-31"),as.Date("2019-12-31")))+
  theme(axis.text.x=element_text(angle=30, hjust=1))
  
  

salmon_graph


```

**Figure 2: Times series plot of daily Steelhead Salmon passage across the Bonneville Dam (1940-2019)**. The times series plots shows a seasonal pattern yearly. There might be a broader cyclical trend which needs further exploration. *Data used with permission from Columbia Basin Research.*

```{r, include=FALSE}


salmon_ts <- as_tsibble(salmon_date, index = yy_mm_dd)


#salmon_ts %>% autoplot(value)



```


```{r fig.align='center'}


# group by year and month and find the total counts
salmon_date2<- salmon_date %>% 
  group_by(year, month) %>% 
  summarise(
    value=sum(value, na.rm= TRUE))

#rename the columns
salmon_date3<-salmon_date2 %>% 
  rename("Year"=year) %>% 
  rename("Month"=month)


salmon_date3$Month = factor(salmon_date3$Month, levels = month.abb)

options(scipen = 999) #getting rid of scientific notation

ggplot(data = salmon_date3, aes(x = Month, y = value, group = Year)) +
  geom_line(aes(color = Year))+
  theme_bw()+
  labs(y="Monthly Total Count")+
  scale_colour_gradient(low = "salmon1", high = "salmon4")+
  theme(text = element_text(size=12))




```

**Figure 3: Seasonplot of monthly total steelhead Salmon passage across the Bonneville Dam over the years**. Colors indicate the years from 1940-2019 going from lighter to darker. The seasonplot shows that there is a seasonal pattern in salmon passage. It starts to increase around May-June, peaks around August every year. *Data used with permission from Columbia Basin Research.*



```{r fig.align='center'}

# use index by function and find annual sums
salmon_annual <- salmon_ts %>% 
  index_by(Year = ~year(.)) %>% 
  summarize(
    total = sum(value,na.rm = TRUE)
  )

ggplot(data = salmon_annual, aes(x = Year, y = total)) +
  geom_line(color = "salmon4") +
  geom_smooth(color = "darkorchid4",
              size = 0.2,
              linetype = "dashed",
              fill = "darkorchid2",
              alpha = 0.1) +
  theme_bw()+
  scale_x_continuous(breaks = seq(1940,2020, by = 5))+
  labs(y="Annual Steelhead Salmon Passage")+
  theme(axis.text.x=element_text(angle=30, hjust=1))+
  theme(text = element_text(size=12))
  
  


```


**Figure 4: Total Annual Steelhead Salmon Passage(1940-2019)**. Smoothed dashed lines are added to observe the overall trends over the years. It is unclear whether there is a broader cyclical trend in Steelhead Salmon passage. Likewise, it is hard to talk about a bigger general trend. *Data used with permission from Columbia Basin Research.*


**REFERENCES**

**[1]** : Columbia River DART, Columbia Basin Research, University of Washington. (2019). Adult Passage Graphics & Text. Available from http://www.cbr.washington.edu/dart/query/adult_graph_text