---
title: Spatial Data Analysis
author: 'Ilayda Dinc'
date: '2020-02-19'
slug: external-project
categories: []
tags:
  - Academic
  - R Markdown
  - Spatial Data Analysis
subtitle: 'Map of Maui Land Use and Watersheds'
summary: 'This project aims to visualize the watersheds and main types of land covers on Maui island.'
authors: [Ilayda Dinc]
lastmod: '2020-02-19T15:02:39-08:00'
featured: no
image:
  caption: 'Maui View from Poli Poli'
  focal_point: ''
  preview_only: no
projects: []
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include=FALSE}
## Attach packages:
library(tidyverse)
library(sf)
library(tmap)
library(leaflet)
library(ggthemes)
library(ggspatial)
library(ggplot2)
library(paletteer)
```


Maui is the second largest Hawaiian island with the surface area of 727.2 square miles [1]. This project aims to visualize the watersheds and main types of land covers on the island.

```{r}
#reading in the data

hawaii_land <- read_sf(dsn = ".", layer = "Land_Use_Land_Cover_LULC") %>% 
  dplyr::select(landcover) %>% 
  filter(!str_detect(landcover, pattern = "0"))

#filtering for main land use types
hawaii_land2<- hawaii_land %>% 
  filter(landcover %in% c("Cropland and Pasture","Residential","Evergreen Forest Land", "Shrub and Brush Rangeland"))
  
#unique(hawaii_land$landcover)
  
hawaii_water <- read_sf(dsn = ".", layer = "Watersheds")

# checking crs 
#st_crs(hawaii_land)  #4326

#st_crs(hawaii_water) #4326

```

```{r, message=FALSE, warning=FALSE,fig.align='center', fig.height=8, fig.width=8}

# grpahing watersheds and land use together
ggplot(hawaii_land)+
  geom_sf(color = "black",
          size = 0.1) + 
  geom_sf(data = hawaii_land2, 
          aes(fill = landcover), color = "NA") +
  scale_fill_manual(values=c("darkgoldenrod2","darkgreen","darkorchid4","palegreen3"),name="Land Cover")+
   geom_sf(data = hawaii_water, 
          color = "dodgerblue4", 
          alpha = 0.3,
          size = 0.15,
          color = "NA") + 
  coord_sf(datum = NA) + # Removes graticules
  geom_point(aes(x = -156.3319, y = 20.7984), size = 2)+
  geom_text(x = -156.3319, y = 20.82, label = "Maui", size=7)+ 
  coord_sf(xlim = c(-155.98,-156.68), ylim = c(20.45,21.1)) +
  theme_bw()+
  labs(x = "Longitude", y = "Latitude")+
  theme(
    legend.position = "right",
    legend.justification = c("right"),
    legend.box.just = "right",
    legend.margin = margin(3, 3, 3, 3),
    legend.text = element_text(size = 7)
    )+
   annotation_scale(location = "br", width_hint = 0.5) +
    annotation_north_arrow(location = "br", which_north = "true", 
        pad_x = unit(0.3, "in"), pad_y = unit(0.3, "in"),
        style = north_arrow_fancy_orienteering)
```
**Figure 2: Map of main land use types and the watersheds on the island of Maui**. The most prominent land use types are Cropland and Pasture, Evergreen Forest Land, and Shrub and Brush Rangeland. Evergreen Forestland is mostly clustred around the watersheds. *Data used with permission from Hawaii Statewide GIS Program.*



**REFERENCES**

**[1]** : "Table 5.08 - Land Area of Islands: 2000", 2004 State of Hawaii Data Book (PDF), State of Hawaii, 2004, retrieved 2007-07-23.




